Methodology of Creating SVM Kernels from Scratch Using Python and NumPy

### 2023-12-15 Paper/paper.tex
Update

\begin{abstract}
This paper presents a methodology for creating Support Vector Machine (SVM) kernels from scratch using Python and NumPy. 
We discuss the implementation of linear, sigmoid, polynomial, and radial basis function (RBF) kernels in a binary and multiclass SVM.
These kernels are tested on E. coli data and compared to the results of the scikit-learn SVM implementation.
The results show that the implemented kernels often yield better accuracy than the scikit-learn implementation; at the
cost of increased training time. Kernel training times are ran multiple times and averaged to provide a more accurate
metric since training times are low and have a high variance.
\end{abstract}

\section{Introduction}
In this section, provide an introduction to SVMs, their applications, and the importance of
kernels in SVMs.

@@

\section{Methodology}
In this section, describe the methodology used to create the SVM kernels from scratch.

@@

\subsection{Binary SVM}
Discuss the implementation of the Binary SVM class, including the implementation of the different kernels.

@@

\section{Results and Discussion}
\subsection{Training Time and Accuracy}
After implimenting the SVM kernels for both binary and multiclass classification, we tested the kernels on the E. coli data set.
Since training times were low and had a higher variance, we ran each kernel 10 times and averaged the results.
This allows for a more accurate representation of the training time for each kernel to prevent outliers from skewing the results.
The results of the training times and accuracy can be seen in Table \ref{tab:classifier_performance}.
\begin{table}[h]
    \centering
    \caption{Classifier Performance}
    \label{tab:classifier_performance}
    \begin{tabular}{cccccc}
        \toprule
        \textbf{Classifier} & \textbf{Implementation} & \textbf{Kernel} & \textbf{Avg Accuracy} & \textbf{Avg Runtime} \\
        \midrule
        Binary & sklearn & linear & 1.00000 & 0.00052 \\
        Binary & custom & linear & 0.99318 & 0.00795 \\
        Binary & sklearn & sigmoid & 0.68180 & 0.00066 \\
        Binary & custom & sigmoid & 0.99546 & 0.00757 \\
        Binary & sklearn & rbf & 1.00000 & 0.00054 \\
        Binary & custom & rbf & 0.99546 & 0.00977 \\
        Binary & sklearn & poly & 1.00000 & 0.00040 \\
        Binary & custom & poly & 0.97498 & 0.00929 \\
        Multi & sklearn & linear & 0.77940 & 0.00096 \\
        Multi & custom & linear & 0.81468 & 0.15420 \\
        Multi & sklearn & sigmoid & 0.47060 & 0.00183 \\
        Multi & custom & sigmoid & 0.82497 & 0.15641 \\
        Multi & sklearn & rbf & 0.75000 & 0.00156 \\
        Multi & custom & rbf & 0.82350 & 0.19177 \\
        Multi & sklearn & poly & 0.76470 & 0.00111 \\
        Multi & custom & poly & 0.83967 & 0.36326 \\
        \bottomrule
    \end{tabular}
    \vspace{1em}  % Add some vertical space for the note
    \begin{tabular}{p{0.9\textwidth}}  % Create a new table for the note
        \multicolumn{1}{l}{\textit{Note:}} These times were calculated with 10 runs for each kernel and averaged. \\
    \end{tabular}
\end{table}
\FloatBarrier 
I think that the results are interesting because the custom implementation of the kernels often outperformed the scikit-learn implementation.
However, the custom implementation of the kernels took significantly longer to train. Since we used Stochastic Gradient Descent (SGD) to train the SVMs, 
the convergence can take a long time. In Sklearn, they use LibSVM which is a C++ library that is much faster than our custom implementation.
Although the multi-class SVMs had lower accuracy than the binary SVMs, the training times were much higher. This makes sense because the multi-class SVMs
are just a combination of binary SVMs and the decision boundaries start to become more complex and harder to fit accurately.
\FloatBarrier 
\subsection{Decision Boundaries}
The decision boundaries for the binary and multi-class SVMs can be seen in Figures 1 and 2.
In Figure 1, the decision boundaries for the binary SVMs are shown. We can see how each kernel affects the decision boundary.
Although these examples are intuitive and easy to understand, the data was reduced to 2 dimensions using Principal Component Analysis (PCA).
We decided to use PCA to reduce the dimensionality of the data because we knew that the training data would retain more information than 
regular feature selection. By implementing feature extraction, we were able to reduce the dimensionality of the data while retaining more information.
This allows us to show the decision boundaries in 2 dimensions while still retaining the accuracy of the SVMs.

>>>>

\begin{abstract}
This paper presents a methodology for creating Support Vector Machine (SVM) kernels from scratch using Python and NumPy. 
We discuss the implementation of linear, sigmoid, polynomial, and radial basis function (RBF) kernels in a binary and multiclass SVM.
These kernels are tested on E. coli data and compared to the results of the scikit-learn SVM implementation.
The goal of the E. coli dataset is to predict the localization sites of proteins inside the cell. 
The results show that the implemented kernels often yield better accuracy than the scikit-learn implementation; at the
cost of increased training time. Kernel training times are ran multiple times and averaged to provide a more accurate
metric since training times are low and have a high variance.
\end{abstract}

\section{Introduction}

@@

\section{Methodology}

@@

\subsection{Binary SVM}

@@

\section{Data Overview}
The data is composed of seven predictors of type float, with the last column being the localization site class label.
There were 336 total observations in the data set, and the class labels were not evenly distributed.
As you can see below, the class labels \textit{cp} and \textit{im} make up the majority of the data set.
We used these two classes as the positive and negative classes for the binary SVMs as they were the most common.
\begin{table}[h]
    \centering
    \caption{Class Frequency}
    \label{tab:class_frequency}
    \begin{tabular}{cc}
        \toprule
        \textbf{Class} & \textbf{Frequency} \\
        \midrule
        cp & 143 \\
        im & 77 \\
        pp & 52 \\
        imU & 35 \\
        om & 20 \\
        omL & 5 \\
        imS & 2 \\
        imL & 2 \\
        \bottomrule
    \end{tabular}
\end{table}
For a test/train split, we used 80\% of the data for training and 20\% for testing.
This resulted in 268 training observations for the multi-class SVMs and 176 training observations for the binary SVMs.
\section{Results and Discussion}
\subsection{Training Time and Accuracy}
After implimenting the SVM kernels for both binary and multiclass classification, we tested the kernels on the E. coli data set.
Since training times were low and had a higher variance, we ran each kernel 10 times and averaged the results.
This allows for a more accurate representation of the training time for each kernel to prevent outliers from skewing the results.
The results of the training times and accuracy can be seen in Table \ref{tab:classifier_performance}.
\begin{table}[h]
    \centering
    \caption{Classifier Performance}
    \label{tab:classifier_performance}
    \begin{tabular}{cccccc}
        \toprule
        \textbf{Classifier} & \textbf{Implementation} & \textbf{Kernel} & \textbf{Avg Accuracy} & \textbf{Avg Runtime} \\
        \midrule
        Binary & sklearn & linear & 1.00000 & 0.00052 \\
        Binary & custom & linear & 0.99318 & 0.00795 \\
        Binary & sklearn & sigmoid & 0.68180 & 0.00066 \\
        Binary & custom & sigmoid & 0.99546 & 0.00757 \\
        Binary & sklearn & rbf & 1.00000 & 0.00054 \\
        Binary & custom & rbf & 0.99546 & 0.00977 \\
        Binary & sklearn & poly & 1.00000 & 0.00040 \\
        Binary & custom & poly & 0.97498 & 0.00929 \\
        Multi & sklearn & linear & 0.77940 & 0.00096 \\
        Multi & custom & linear & 0.81468 & 0.15420 \\
        Multi & sklearn & sigmoid & 0.47060 & 0.00183 \\
        Multi & custom & sigmoid & 0.82497 & 0.15641 \\
        Multi & sklearn & rbf & 0.75000 & 0.00156 \\
        Multi & custom & rbf & 0.82350 & 0.19177 \\
        Multi & sklearn & poly & 0.76470 & 0.00111 \\
        Multi & custom & poly & 0.83967 & 0.36326 \\
        \bottomrule
    \end{tabular}
    \vspace{1em}  % Add some vertical space for the note
    \begin{tabular}{p{0.9\textwidth}}  % Create a new table for the note
        \multicolumn{1}{l}{\textit{Note:}} These times were calculated with 10 runs for each kernel and averaged. \\
    \end{tabular}
\end{table}
\FloatBarrier 
I think that the results are interesting because the custom implementation of the kernels often outperformed the scikit-learn implementation.
However, the custom implementation of the kernels took significantly longer to train. Since we used Stochastic Gradient Descent (SGD) to train the SVMs, 
the convergence can take a long time. In Sklearn, they use LibSVM which is a C++ library that is much faster than our custom implementation.
Although the multi-class SVMs had lower accuracy than the binary SVMs, the training times were much higher. This makes sense because the multi-class SVMs
are just a combination of binary SVMs and the decision boundaries start to become more complex and harder to fit accurately.
\FloatBarrier 
\subsection{Decision Boundaries}
The decision boundaries for the binary and multi-class SVMs can be seen in Figures 1 and 2.
In Figure 1, the decision boundaries for the binary SVMs are shown. We can see how each kernel affects the decision boundary.
Although these examples are intuitive and easy to understand, the data was reduced to 2 dimensions using Principal Component Analysis (PCA).
We decided to use PCA to reduce the dimensionality of the data because we knew that the training data would retain more information than 
regular feature selection. By implementing feature extraction, we were able to reduce the dimensionality of the data while retaining more information.
This allows us to show the decision boundaries in 2 dimensions while still retaining the accuracy of the SVMs.
